---
title: "cLHS + Random Forest Optimization"
author: "soilsampling Package Authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cLHS + Random Forest Optimization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

Conditioned Latin hypercube sampling (cLHS) gives a design that is *environmentally representative*, but representativeness is not the same as *predictive accuracy*. Wadoux et al. (2019) showed that a cLHS baseline can be further improved for a specific prediction task by directly optimizing the sample locations against a Random Forest's cross-validated prediction error.

This package implements that idea as a two-stage workflow:

1. **cLHS baseline**: generate a representative starting design (`ss_clhs_sample()`)
2. **Simulated annealing refinement**: iteratively swap sample points to minimize Random Forest MSE (`ss_rf_optimize()`)

`ss_rf_size()` runs both stages end-to-end and produces comparison plots.

## Theoretical Background

### The Objective Function

At each candidate design, `ss_rf_mse()` fits a Random Forest with $k$-fold cross-validation, predicting one covariate (the "response") from the others, and returns the mean squared error averaged across folds. Lower MSE means the design's points are more informative for predicting that variable from the rest.

### Simulated Annealing

`ss_rf_optimize()` starts from the cLHS baseline and repeats, for `n_iterations`:

1. Replace one randomly chosen sample point with a randomly chosen point not currently in the design
2. Compute the candidate design's RF MSE
3. **Accept** the swap if it improves MSE, or **probabilistically accept** a worse swap with probability $\exp(-\Delta / T)$ (the Metropolis criterion)
4. Cool the temperature: $T \leftarrow T \times \text{cooling\_rate}$

Early iterations (high temperature) tolerate worse moves, allowing the search to escape local optima; later iterations (low temperature) only accept improvements, converging to a refined design. The best design seen at any point is tracked and returned, regardless of where the search ends up.

## Basic Usage

As in the KL sample size vignette, we build a small synthetic covariate stack so the example runs quickly.

```{r message=FALSE}
library(soilsampling)
library(terra)

set.seed(7)

nr <- 30
nc <- 30

r <- rast(nrows = nr, ncols = nc, xmin = 0, xmax = 100, ymin = 0, ymax = 100, nlyrs = 3)
names(r) <- c("dem", "slope", "ndvi")

xy <- xyFromCell(r, seq_len(ncell(r)))
values(r)[, "dem"] <- 100 + 20 * sin(xy[, 1] / 15) + rnorm(ncell(r), 0, 3)
values(r)[, "slope"] <- abs(cos(xy[, 2] / 12) * 8 + rnorm(ncell(r), 0, 1))
values(r)[, "ndvi"] <- pmin(pmax(0.5 + 0.3 * cos(xy[, 1] / 20 + xy[, 2] / 25) +
  rnorm(ncell(r), 0, 0.05), -1), 1)

r
```

### Running the End-to-End Workflow

```{r}
result <- ss_rf_size(
  r,
  n_samples = 20,
  n_iterations = 30,
  seed = 123
)

result$comparison_table
result$improvement
```

`result$improvement` is the percent reduction in cross-validated MSE achieved by the RF-optimized design relative to the cLHS baseline.

### Visualizing the Comparison

```{r}
result$plots$clhs_plot
result$plots$rf_plot
```

## Function Walkthrough

### cLHS Baseline Alone

```{r}
baseline <- ss_clhs_sample(r, n_samples = 20, seed = 123)
head(baseline)
```

### Evaluating a Design's MSE Directly

```{r}
ss_rf_mse(baseline)
```

By default, the third covariate column is used as the response (predicted from the rest); pass `target_var` to choose a different one.

```{r}
ss_rf_mse(baseline, target_var = "ndvi")
```

### Running the Optimizer on Its Own

```{r}
opt <- ss_rf_optimize(
  r,
  n_samples = 20,
  n_iterations = 30,
  seed = 123
)

opt$initial_mse
opt$final_mse
opt$improvement
```

## Choosing Parameters

- **`n_iterations`**: more iterations give the search more chances to improve, at a roughly linear cost in runtime (one RF fit per iteration). This vignette uses 30 to stay fast; real analyses typically use several hundred.
- **`temperature` / `cooling_rate`**: higher initial temperature and slower cooling (`cooling_rate` closer to 1) explore more broadly before converging, at the cost of more iterations needed to settle. The defaults (`1000` / `0.95`) work well for designs of a few dozen to a few hundred points.
- **`target_var`**: choose the covariate that best represents your actual prediction target (e.g. an observed soil property in real data), not necessarily the default third column.
- **`seed`**: the cLHS baseline uses `seed`; the annealing loop uses `seed + 1`, so both stages are reproducible independently.

## Practical Workflow

```{r eval=FALSE}
library(soilsampling)
library(terra)

predictors <- rast("data/predictors.tif")

result <- ss_rf_size(
  predictors,
  n_samples = 100,
  n_iterations = 500,
  seed = 123,
  output_dir = "outputs"
)

result$comparison_table
result$improvement

# outputs/ now contains:
#   clhs_sample_locations.csv, rf_optimized_locations.csv,
#   sampling_comparison_table.csv, sampling_comparison_plots.png
```

## When to Use cLHS + RF Optimization

✅ **Use it when:**

- You have a specific covariate (or measured property) you want to predict well, not just represent
- A Random Forest (or similar model) is a realistic stand-in for your eventual prediction workflow
- You can afford the extra computation of iterative RF fitting on top of cLHS

❌ **Don't rely on it when:**

- You need probability-based sampling for design-based (not model-based) inference
- Your prediction target isn't known yet, or differs substantially from the covariates used for optimization
- The candidate pool is very large and RF fitting per iteration becomes prohibitively slow — consider a coarser candidate grid first

## References

- Wadoux, A.M.J-C., Brus, D.J., and Heuvelink, G.B.M. (2019). Sampling design optimization for soil mapping with random forest. *Geoderma* 355, 113913. DOI: [10.1016/j.geoderma.2019.113913](https://doi.org/10.1016/j.geoderma.2019.113913)

- Minasny, B., and McBratney, A.B. (2006). A conditioned Latin hypercube method for sampling in the presence of ancillary information. *Computers & Geosciences* 32(9), 1378-1388. DOI: [10.1016/j.cageo.2005.12.009](https://doi.org/10.1016/j.cageo.2005.12.009)

- Breiman, L. (2001). Random forests. *Machine Learning* 45(1), 5-32. DOI: [10.1023/A:1010933404324](https://doi.org/10.1023/A:1010933404324)

- Kirkpatrick, S., Gelatt, C.D., and Vecchi, M.P. (1983). Optimization by simulated annealing. *Science* 220(4598), 671-680. DOI: [10.1126/science.220.4598.671](https://doi.org/10.1126/science.220.4598.671)
