---
title: "Introduction to soilsampling"
author: "soilsampling Package Authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to soilsampling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

The **soilsampling** package provides methods for designing soil sampling schemes, including:

- **Simple Random Sampling**: Uniform random selection of sampling locations
- **Stratified Random Sampling**: Random sampling within compact geographical strata
- **Spatial Coverage Sampling**: Purposive sampling at stratum centroids for optimal coverage
- **Maxvol Optimal Design**: D-optimal design using maximum volume criterion
- **Composite Sampling**: Sampling from equal-area strata for combined samples

The package uses **sf** for spatial operations and implements pure R algorithms, requiring no Java or additional GIS software.

## Installation

```{r eval=FALSE}
# Install from local source
install.packages(".", repos = NULL, type = "source")

# Or using devtools
devtools::install()
```

## Quick Start

```{r message=FALSE}
library(soilsampling)
library(sf)

# Create a study area
poly <- st_polygon(list(rbind(
  c(0, 0), c(100, 0), c(100, 50), c(0, 50), c(0, 0)
)))
study_area <- st_sf(geometry = st_sfc(poly))
```

### Spatial Coverage Sampling

```{r}
# Set seed for reproducibility
set.seed(42)

# Create spatial coverage sampling design
samples_coverage <- ss_coverage(study_area, n_strata = 25, n_try = 5)

# View summary
print(samples_coverage)

# Plot
ss_plot(samples_coverage)
```

### Stratified Random Sampling

```{r}
# First create strata
strata <- ss_stratify(study_area, n_strata = 25, n_try = 5)

# Take random samples within each stratum
samples_stratified <- ss_stratified(strata, n_per_stratum = 1)

# Plot
ss_plot(strata, samples = samples_stratified)
```

### Simple Random Sampling

```{r}
# Simple random sampling
samples_random <- ss_random(study_area, n = 25)

# Plot
ss_plot_samples(samples_random)
```

## Exporting Results

### As Data Frame

```{r}
# Get coordinates as data frame
coords <- ss_to_data_frame(samples_coverage)
head(coords)
```

### As CSV

```{r eval=FALSE}
# Export to CSV
write.csv(coords, "sampling_points.csv", row.names = FALSE)
```

### As Shapefile or GeoPackage

```{r eval=FALSE}
# Get as sf object
samples_sf <- ss_to_sf(samples_coverage)

# Export to GeoPackage
st_write(samples_sf, "sampling_points.gpkg")

# Export to Shapefile
st_write(samples_sf, "sampling_points.shp")
```

## Working with Real Data

In practice, you would load a study area from a shapefile:

```{r eval=FALSE}
# Read study area from shapefile
study_area <- st_read("path/to/study_area.shp")

# Create sampling design
set.seed(42)
samples <- ss_coverage(study_area, n_strata = 50, n_try = 10)

# Export coordinates
coords <- ss_to_data_frame(samples)
write.csv(coords, "field_sampling_points.csv", row.names = FALSE)
```

## When to Use Each Method

| Method | Best For | Inference Type | Key Features |
|--------|----------|----------------|--------------|
| `ss_coverage()` | Interpolation (kriging) | Model-based | Optimal spatial coverage |
| `ss_stratified()` | Estimating means/totals | Design-based | Valid probability sample |
| `ss_random()` | Simple design-based analysis | Design-based | Unbiased estimates |
| `ss_maxvol()` | Optimal feature coverage | Model-based | D-optimal design |
| `ss_composite()` | Reducing laboratory costs | Design-based | Combined samples |

## Sampling Methods Comparison

### Design-Based vs Model-Based

**Design-based sampling** (e.g., `ss_random()`, `ss_stratified()`):
- Uses probability sampling
- Statistical inference based on sampling design
- Valid for design-based estimation (means, totals)
- No assumptions about spatial autocorrelation required

**Model-based sampling** (e.g., `ss_coverage()`, `ss_maxvol()`):
- Purposive (non-random) sample selection
- Requires spatial model for inference (e.g., kriging)
- Optimal for prediction and interpolation
- More efficient for mapping applications

## Tips for Best Results

1. **Use multiple tries**: Set `n_try = 10` or higher to avoid local optima in k-means algorithms
2. **Set a seed**: Use `set.seed()` for reproducible results
3. **Adjust resolution**: Use `n_cells` parameter to control grid density
4. **Check convergence**: The output includes a `converged` flag
5. **Consider prior information**: Use `prior_points` to incorporate existing sample locations

## Getting Help

For more detailed information on specific methods, see:

- `vignette("spatial-coverage")` - Spatial coverage sampling methods
- `vignette("maxvol-sampling")` - Maxvol optimal design sampling
- `?ss_stratify` - Help on stratification
- `?ss_coverage` - Help on coverage sampling
- `?ss_maxvol` - Help on maxvol sampling

## References

- de Gruijter, J.J., Brus, D.J., Bierkens, M.F.P., and Knotters, M. (2006). *Sampling for Natural Resource Monitoring*. Springer, Berlin.

- Walvoort, D.J.J., Brus, D.J., and de Gruijter, J.J. (2010). An R package for spatial coverage sampling and random sampling from compact geographical strata by k-means. *Computers & Geosciences* 36, 1261-1267.

- Petrovskaia, A., Ryzhakov, G., & Oseledets, I. (2021). Optimal soil sampling design based on the maxvol algorithm. *Geoderma*, 383, 114733.
