---
title: "KL Sample Size Optimization"
author: "soilsampling Package Authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{KL Sample Size Optimization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

Conditioned Latin hypercube sampling (cLHS) is widely used to select environmentally representative sampling locations, but it leaves one question open: **how many samples are enough?** Too few samples fail to represent the environmental variability of the study area; too many waste field and laboratory resources.

Malone et al. (2019) proposed answering this using **Kullback-Leibler (KL) divergence**: run cLHS at a range of candidate sample sizes, measure how well each sample's distribution matches the population's distribution across all covariates, and pick the smallest sample size that reaches a target level of representativeness.

This vignette walks through `ss_kl_size()` and its building blocks: `ss_load_rasters()`, `ss_kl_divergence()`, `ss_kl_optimize()`, and `ss_kl_save_plots()`.

## Theoretical Background

### KL Divergence Between Population and Sample

For each numeric covariate, both the population and a candidate sample are binned into histograms over the same breaks. The KL divergence for that variable is:

$$
KL = \sum_i O_i \log\left(\frac{O_i}{E_i}\right)
$$

where $O_i$ is the sample's density in bin $i$ and $E_i$ is the population's density in the same bin. `ss_kl_divergence()` averages this across all numeric covariates, giving a single divergence score per sample: lower is better (more representative).

### From Divergence to Optimal Sample Size

`ss_kl_optimize()` repeats this at a sequence of sample sizes (`min_samples` to `max_samples`, in `step_size` increments), running `n_replicates` independent cLHS draws at each size. The mean KL divergence per size is then fit to an exponential decay curve:

$$
KL(n) = b_1 e^{-kn} + b_0
$$

The **optimal sample size** is the smallest $n$ for which the cumulative proportion of achievable improvement,

$$
\frac{\max(KL) - KL(n)}{\max(KL) - \min(KL)},
$$

reaches `probability_threshold` (default `0.95`) — i.e. the point of diminishing returns.

## Basic Usage

`ss_kl_size()` accepts a raster path, an already-loaded `SpatRaster`, or a plain population data frame. For this vignette we build a small synthetic covariate stack so the example runs quickly and needs no external files.

```{r message=FALSE}
library(soilsampling)
library(terra)

set.seed(42)

nr <- 40
nc <- 40

# Synthetic terrain-like covariates with spatial structure + noise
r <- rast(nrows = nr, ncols = nc, xmin = 0, xmax = 100, ymin = 0, ymax = 100, nlyrs = 3)
names(r) <- c("dem", "slope", "ndvi")

xy <- xyFromCell(r, seq_len(ncell(r)))
values(r)[, "dem"] <- 100 + 20 * sin(xy[, 1] / 15) + rnorm(ncell(r), 0, 3)
values(r)[, "slope"] <- abs(cos(xy[, 2] / 12) * 8 + rnorm(ncell(r), 0, 1))
values(r)[, "ndvi"] <- pmin(pmax(0.5 + 0.3 * cos(xy[, 1] / 20 + xy[, 2] / 25) +
  rnorm(ncell(r), 0, 0.05), -1), 1)

r
```

### Running the End-to-End Workflow

```{r}
result <- ss_kl_size(
  r,
  min_samples = 10,
  max_samples = 60,
  step_size = 10,
  n_replicates = 3,
  n_bins = 15
)

result$optimal_sample_size
result$summary_results
```

`result$summary_results` shows the mean and standard deviation of KL divergence at each tested sample size — you should see it decrease as sample size grows, then flatten out.

### Visualizing the Result

```{r}
result$plot_kl
result$plot_cdf
```

`plot_kl` shows the raw KL divergence values with the fitted exponential decay curve; `plot_cdf` shows the cumulative-improvement curve used to pick `optimal_sample_size`, with the threshold and chosen size marked.

## Function Walkthrough

### Loading Predictor Rasters

```{r eval=FALSE}
# Single multi-layer file
predictors <- ss_load_rasters("data/predictors.tif")

# Or a directory of single-layer .tif files
predictors <- ss_load_rasters("data/covariates/")
```

### Computing KL Divergence Directly

If you already have a population and a candidate sample as data frames, you can skip the raster/cLHS machinery entirely:

```{r}
population <- as.data.frame(r, na.rm = TRUE)
candidate_sample <- population[sample(nrow(population), 30), ]

ss_kl_divergence(population, candidate_sample, n_bins = 15)
```

### Running the Optimizer on a Data Frame

`ss_kl_optimize()` is the engine behind `ss_kl_size()`. Use it directly when you already have a population data frame (e.g. from your own raster-to-data-frame pipeline):

```{r}
opt <- ss_kl_optimize(
  population,
  min_samples = 10,
  max_samples = 60,
  step_size = 10,
  n_replicates = 3,
  n_bins = 15
)

opt$optimal_sample_size
```

### Saving Plots Independently

`ss_kl_size(..., output_dir = ...)` writes CSVs and plots together. To save just the plots from a result you already have in memory (e.g. under a different naming scheme), use `ss_kl_save_plots()`:

```{r eval=FALSE}
ss_kl_save_plots(result, output_dir = "my_outputs", prefix = "clhs_kl")
```

`output_dir` has no default: the function writes files, so the caller
always chooses where they land.

## Choosing Parameters

- **`step_size`**: smaller steps give a finer-grained curve but multiply the number of cLHS runs. Start coarse (e.g. 20-50 for a few hundred candidate samples) and refine around the optimum if needed.
- **`n_replicates`**: cLHS is stochastic, so each sample size needs multiple draws to estimate a stable mean KL divergence. 10 is a reasonable default for real analyses; this vignette uses 3 to keep runtime short.
- **`n_bins`**: number of histogram bins per covariate. Too few bins blur real differences; too many make the divergence estimate noisy with small samples.
- **`probability_threshold`**: how much of the achievable improvement you require before stopping. `0.95` (default) is a common choice; raise it if you can afford a larger sample, lower it if resources are tight.

## Practical Workflow

```{r eval=FALSE}
library(soilsampling)

result <- ss_kl_size(
  "data/predictors.tif",
  min_samples = 10,
  max_samples = 500,
  step_size = 10,
  n_replicates = 10,
  probability_threshold = 0.95,
  output_dir = "outputs"
)

result$optimal_sample_size

# outputs/ now contains:
#   kl_raw_results.csv, kl_summary_results.csv, kl_fitted_curve.csv,
#   kl_divergence_vs_sample_size.png, kl_cdf_threshold.png
```

## When to Use KL Sample Size Optimization

✅ **Use it when:**

- You're planning a cLHS design and need to justify the sample size
- You want a data-driven stopping point instead of a fixed round number
- Field/laboratory costs make oversampling expensive

❌ **Don't rely on it when:**

- Your sample size is fixed by budget or logistics regardless of the analysis
- You have fewer than ~4 candidate sample sizes to test (the exponential fit needs more points to converge)

## References

- Malone, B.P., Minasny, B., and Brungard, C. (2019). Some methods to improve the utility of conditioned Latin hypercube sampling. *PeerJ* 7, e6451. DOI: [10.7717/peerj.6451](https://doi.org/10.7717/peerj.6451)

- Minasny, B., and McBratney, A.B. (2006). A conditioned Latin hypercube method for sampling in the presence of ancillary information. *Computers & Geosciences* 32(9), 1378-1388. DOI: [10.1016/j.cageo.2005.12.009](https://doi.org/10.1016/j.cageo.2005.12.009)

- Saurette, D.D., Heck, R.J., Gillespie, A.W., Berg, A.A., and Biswas, A. (2023). Divergence metrics for determining optimal training sample size in digital soil mapping. *Geoderma* 436, 116553. DOI: [10.1016/j.geoderma.2023.116553](https://doi.org/10.1016/j.geoderma.2023.116553)
